On the properties of discs around accreting brown dwarfs

### 2010-07-12 BDD_submit/paper.tex
First response to referee very minor changes
git-svn-id: https://repository.astro.ex.ac.uk/torus/trunk@2364 25464f21-6099-402c-b30a-640f1d40d38b

The interpretation of the spectral energy distributions (SEDs) of
pre-main-sequence (PMS) stars involves trying to distinguish the
various contributions to the continuum from the hot spots at the base
of the accretion flow, the photospheric flux, and the near-IR flux
from the dusty inner disc. The complexity of the interplay between
these contributions is excerbated by other, geometrical, effects such
as the inclination (which changes the projected area of the inner disc
wall visible to the observer) and the outer disc structure (which may
obsure the inner disc for high inclincations)
\citep{walker_2004,tannirkulam_2007}.

@@

Comparison of accretion rates across the pre-main-sequence (PMS) mass
spectrum has indicated that the accretion rate is strongly correlated
with PMS mass, \citep{muzerolle_2003,natta_2004,natta_2006}, with an
approximate form $\dot{M}_{\rm acc}\propto M^2_*$. Since in the
canonical picture the accretion rate is driven by the disc viscosity,
and should be indepedent of the mass of the central object, this
correlation is somewhat surprising. There is some danger that the
correlation is the result of, or at least strengthened by, the
presence of observational biases \citep{clarke_2006}. At the high-mass
(CTTS) end of the mass spectrum the lowest accretion-rates cannot be
measured via continuum methods since the excess is too small in
contrast with the photopsheric emission. The emission lines will also
be weak, and indeed the H$\alpha$ equivalent width (EW) may be less
than the 10\AA\ that traditionally demarcates classical from
weak-lined T Tauri stars. Such objects should show doppler-broadened
profiles, but the line wings will be weak, and the presence of
underlying H$\alpha$ absorption may become dominant. Thus there may
well be a population of low-accretion-rate CTTS which are current
missing from the surveys.

There are also risks of observational biases at the low-mass end of
the correlation. Radiative-transfer modelling of H$\alpha$ emission
from accreting BDs requires high temperatures ($>$\, 10\,kK) in the
accretion funnels in order to recover the level of emission that is
observed \citep{muzerolle_2003,natta_2004,natta_2006}. Although
cooling rate arguments can be invoked to explain the presence of such
high temperatures, the line ratios of Pa$\beta$ and Br$\gamma$
indicate that the hydrodgen emission may not arise in the funnel flows
\citep{gatti_2006}. However, recently some continuum measurements of
accreting BDs have been conducted \citep{herczeg_2009}, and these
broadly support the H$\alpha$ rates.

@@

In order to model an accreting BDD system we must first model the
underlying photospheric flux. We have adopted a BD stellar interior
and atmospheric model grid and have then constructed the total
photospheric flux for any input value of stellar age and mass by
interpolating for surface gravity ($\log g$), effective temperature
($T_{\rm eff}$), radius ($R_*/R_{\odot}$) and luminosity
($L_*/L_{\odot}$). These values were then used to interpolate
atmospheric spectra for flux (ergs s$^{-1}$cm$^{-2} {\rm \AA}^{-1}$)
from 1200 to 2$\times$10$^7$ \AA. The spectra were subsequently
resampled onto 200 logarithmically spaced points. Careful inspection
ensured that no spectral features were removed during resampling. The
stellar interior models used for this study are the `DUSTY00' models
of \cite{chabrier_2000} combined with the `AMES-Dusty', atmospheric
models of \cite{chabrier_2000}, which are all available
online\footnote{http://perso.ens-lyon.fr/france.allard/}. For our
$T_{\rm eff}$ range of $\approx $3000\,K$<T_{\rm eff}<$1600\,K the
AMES-Dusty atmospheres are the most applicable (2700\,K$>T_{\rm
  eff}>$1700\,K). We did try including dynamic application of
atmospheres based on the derived $T_{\rm eff}$, i.e. using AMES-Cond
for $T_{\rm eff}<$1700\,K, but this resulted in large discontinuities
between the model atmospheres and resulting spectra. Since this only
affects stars at the very edge of our temperature range, i.e. for the
oldest and lowest mass objects (for the AMES-Cond case), we have
adopted the AMES-Dusty models throughout.

@@

Finally, a blackbody flux distribution is generated at $T_{\rm acc}$
and added onto the intrinsic stellar photospheric flux. In general one
would expect this to be an overestimate of the accretion flux, as for
pre-MS BDs large convective zones are expected on the stellar surface,
and some of the accretion energy may act to further drive these
convective currents, meaning flux is lost. It is worth noting however
that observationally UV excesses are often used to recreate and then
subtract an assumed accretion flux using a blackbody flux curve, which
is essentially the reverse of this method.

@@

As our models assume magnetic-truncation at the co-rotation radius, we
have implicitly generated an inner hole. This also means that the disc
will have an inner wall at this radius. Evidence for inner walls in
circumstellar discs is apparent from the SEDs of disc systems, where a
peak in emission is found between 2 and 3\,$\mu$m. The temperatures
reached by such inner walls are expected to generate thermal flux
contributions within this wavelength range \citep{dullemond_2001}. The
circumstellar disc receives direct photospheric radiation and is
strongly heated. This leads to an increased scaleheight, or `puffing
up' of the inner disc rim, and subsequent shadowing of the region
immediately beyond it \citep{dullemond_2001}. The density-dependent
nature of the dust sublimation temperature \citep{pollack_1994} may
also play a role in shaping the inner rim
\citep{isella_2005,tannirkulam_2007}\footnote{Although we recognise
  that the inner hole in protostellar discs may be the result of giant
  planet formation or photoevaporation \citep{dahm_2009,najita_2007},
  these two mechanisms should result in a negligible mass-accretion
  rate \citep{dahm_2009} and therefore would represent a distinct
  evolutionary state of brown dwarf disc (BDD) sytems to that of the
  CTTS-like phase we are considering here.}.

@@

Our treatment of dust sublimation is similar to that detailed in
\cite{tannirkulam_2007}, but with some enhancements to ensure a swift
convergence of the sublimated rim. The dust sublimation proceeds as
follows: An initial temperature distribution is found for the
optically-thin limit by setting the global dust-to-gas ratio to a tiny
value. Subsequent radiative equilibrium iterations are performed using
the adopted dust-to-gas ratio of 0.01, but limiting the maximum
optical depth across a given cell to $\tau_{\rm max}$. Cells whose
temperature exceeds the local dust sublimation temperature have their
dust-to-gas ratio set to zero. Radiative equilibrium iterations and
sublimation sweeps are performed at $\tau_{\max} = $0.1, 1, and 10,
with a final iteration of $\tau_{\max} = \infty$. Adequately solving
the radiative-equilibrium necessiates resolving the disc's effective
photosphere, and adaptive mesh refinement is used to split the grid at
the optically-thin/optically-thick boundary in order that the maximum
cell size at this boundary is $<1$ at a wavelength of 5500\,\AA. Once
a self-consistent sublimated rim has been determined the equation of
hydrostatic equilibrium is solved throughout the disc, and the
sublimation iterations are restarted (the change in the density
structure from the hydrostatical equilibrium step naturally feeds back
into the shape of the inner rim). Five hydrostatic equilibrium steps
are performed to ensure convergence, although a stable disc structure
is normally found after three such iterations. For a more
sophisticated, and chemically more complex, study of dust sublimation,
albeit for higher mass objects see \cite{kama_2009}.

@@

Accretion rates derived for pre-MS stars are of order \logmdot = $-6$
to $-11$ \citep{natta_2006}, with the largest accretion rates found in
so-called FU Orionis type objects. For the more typical accretion
rates \citep[$\dot{M}=$10$^{-11}$ to 10$^{-8.9} M_{\odot}yr^{-1}$, for
  TTS,][]{dahm_2009}, several studies have now suggested that the
accretion rate is strongly correlated with the mass of the central
star. This relationship was perhaps first suggested by
\cite{muzerolle_2003} using various accretion diagnostics. Later,
\citep{muzerolle_2005} derived a relationship of approximately
$\dot{M}\propto M_*^2$. Further evidence was put forward by
\cite{natta_2004}, where accretion rates as low as 5$\times$
10$^{-12} {\rm M}_{\odot} {\rm yr}^{-1}$ were found for BD stars. More recently,
even lower accretion rates of $\approx$10$^{-13} {\rm M}_{\odot} {\rm yr}^{-1}$,
have been derived for BDs by \cite{herczeg_2009}.  Further support for
a dependence of accretion rate on stellar mass was apparent in the
significantly more homogeneous dataset of
\cite{natta_2006}. \cite{natta_2006} analysed a set of accretion rates
and masses derived for pre-MS stars in $\rho$ Ophiuchi and compared
these results to stars in Taurus. They found that the accretion rate
scales with central object mass into the BD regime, although with
significant scatter.

@@

For further analysis, and throughout the rest of this paper, we will
seperate the systems into two groups. These groups are those systems
with \logmdot$\leq-9$ and \logmdot$\geq-9$, classed as typical and
extreme accretors. This distinction is based on the point at which the
systems undergo both significant sublimation ($\Delta R_{\rm
  inner}>1R_*$) and have an accretion luminosity that is comparible to
the photospheric luminosity (see Section~\ref{acc_dominance}). In
order to examine the effect of dust sublimation on our models we have
measured the radius of the inner edge of the dust disc by integrating
from the centre along the midplane until an optical depth of 2/3 (at
5500\AA) is reached. Figure \ref{inner_temp} shows the radial density
distribution of the disc.  In this case the the final inner edge
radius and therefore temperature is no longer strongly correlated with
the rotation rate. Figure \ref{inner_temp} shows the final inner edge
location ($R_{\rm inner}$, $R_*$) against the temperature of the inner
wall ($T_{\rm inner}$, K). The left panel shows the systems designated
as typical accretors and the right panel those with extreme accretion
rates.  For both panels the systems with rotation periods of 0.5 and 5
days are plotted in red and blue respectively. Those systems where the
change in inner radius was greater than 1$R_*$, and therefore classes
as significantly sublimated, are plotted as crosses.

\begin{figure*}
\includegraphics[scale=0.5,angle=90]{./Fig/inner_temp.ps}
   \caption{Figure showing data for the inner edge location ($R_{\rm
       inner}$, $R_*$). The typical accretors are shown in the
     left panel and the extreme accretors in the
     right panel (see text for explanation). In both
       panels separate the systems by period, with those systems with
     rotation periods of 0.5 and 5 days shown as red and blue symbols
     respectively. In addition, systems where $\Delta R_{\rm inner} <
     1R_*$ are plotted ia crosses and $\Delta R_{\rm inner} > 1R_*$ as
     filled circles (this is only achieved by some systems classed as
     extreme accretors).\label{inner_temp}}
\end{figure*}

@@

For our model grid the inner edge shape clearly changes with
increasing flux levels and therefore accretion rate. This, for
individual systems will lead to the reduced dependence of the IR
colours on inclination as the edge moves from a vertical to a convex
boundary as found by \cite{isella_2005}. Practically, for populations
of stars however disentangling this effect from the influences of
various other parameters (such as changing disc scaleheights and
accretion rates) on the IR colours is difficult. Essentially, matching
flux levels between two systems whilst maintinaing a difference in
inner wall shape is impossible.

Therefore, for individual systems changes in the accretion rate and
disc structure lead to significant changes in magnitudes and
colours. The data presented in this section are for an isolated mass
and age. However, the trends described are evident throughout our
model grid for any given subset, and as such are representative. The
scatter or changes in magnitude in the individual systems will change
as a function of the remaining variables, but the dominant input
variables affecting simulated photometry are accretion rate and
inclination. For populations of stars these changes in magnitude and
colour act to scatter or spread a pre-MS BD locus in photometric
space.

@@

Practically, most parameters for young pre-MS stars are derived from
surveys of populations, usually open clusters, using broadband
photometry and subsequently constructed colour-magnitude and
colour-colour diagrams (CMDs and CoCoDs respectively, hereafter).  In
this Section we explore the consequences of our model grid on the
derivation of the primary parameters of age, mass and disc fractions,
from populations. This in turn leads to highlighting selection effects
with, for instance, the mass to accretion rate relation.

@@

As discussed in Section \ref{disc_struct} the inner edge location is
correlated with inner edge temperature (albeit differently for the
typical and extreme accretors). As noted by \cite{meyer_1997} this
could lead to a correlation of IR excess with inner edge position. As
expected however, scatter caused by variations in inclination and disc
structure act to remove this correlation for populations.  The
correlation for an individual set of a systems, i.e. all variables
fixed except rotation rate, between IR colour and rotation rate is
preserved. However, as shown in Section \ref{disc_struct} the changes
caused by accretion rate and inclination angle are the most
significant, and act to remove any correlation between rotation rate
and IR colours for populations. In fact, for our grid the simulated
photometry shows no significant correlation between rotation rate and
IR colours. The work of \cite{meyer_1997} studied flat accreting disc,
therefore, as we have increased variation in disc structure by
applying vertical hydrostatic equilibrium it would be interesting to
see if any correlation appears for analytically defined disc
structures. We have run a set of parrallel models using analytical
disc structures and will publish the results in a subsequent
publication. Overall, as found in \cite{walker_2004}, the dominant
scattering effect for BD disc systems in an optical CMD appears to be
caused by accretion rate and inclination. Therefore for derivation of
parameters we explore the scatter in pre-MS BD locii as a function of
accretion rate and inclination.

@@

The use of pre-MS isochrones for the derivation of single star
parameters is at present not proven to be reliable \citep[see
discussion in][]{mayne_2008}. Practically, therefore, median ages are
derived from populations. Subsequently, derived masses are still
unreliable but at least based on a consistent age. This problem is
being addressed by Bell et al (in prep), where $K$ band photometry and
known eclipsing binaries are being used to refine pre-MS
isochrones. In this section we plot the data for our 1 Myr systems
only and explore the resulting scatters caused by the disc presence
and accretion luminosity.

Figures \ref{age_mass_VVI} and \ref{age_mass_JJK} shows CMDs in $M_V$
\& $(V-I)_0$ and $M_J$ \& $(J-K)_0$ respectively. The left panels of
both figures shows stars classed as typical accretors with accretion
rates of \logmdot = $-$9, $-$10 and $-$11 \& $-$12, shown as blue,
black and red dots respectively in the top panels. The right panels of
Figures \ref{age_mass_VVI} and \ref{age_mass_JJK} show systems with
extreme accretion rates, with \logmdot = $-$6, $-$7 and $-$8 shown as
blue, black and red dots respectively, in the top right panel. The
bottom panels then show the systems seperated into groups by
inclination. These groups are $\theta \leq 48^{\circ}$ as blue dots
(classed as face-on systems), $\theta> 56$ \& $64^{\circ}$ (classed as
the expected systems, as the expectation value of
$cos(\theta)=60^{\circ}$) as black dots and $\theta \geq 71^{\circ}$
(classed as edge-on systems) as red dots (for typical, bottom left,
and extreme, bottom right accretion rates). The left panels then show
naked stars isochrones (created from our grid of simulated photometry)
for 1 Myrs at accretion rates of \logmdot = $-$12 and $-$9, as solid
and dashed lines respectively (with coverage of 10\% and rotation
rates of 5 days). Whereas the left hand panels show naked stars
isochrones of \logmdot = $-$6 and $-$9 as solid and dashed black lines
respectively (with coverage of 10\% and rotation rates of 5 days. The
solid green line, in the top panels, shows the 1 Myr isochrone of
\cite{siess_2000} adjusted to a distance of 250pc and an extinction of
$A_V=2$ mag, simulating a background population of CTTS stars.

\begin{figure*}
  \centering
  \includegraphics[scale=0.6,angle=90]{./Fig/age_mass_VVI.ps}
  \caption{Figure showing CMDs in $M_V$, $(V-I)_0$ for typical
    accretors (\logmdot $\leq$ $-$9) in the left panels, and extreme
    accretors (\logmdot $>-9$) in the right panels. The black dashed
    and solid lines correspond to naked BD isochrones (from our grid)
    with \logmdot\ $-$9 and $-$12 for the left panels (typical
    accretors) and -9 and -6 in the right panels (extreme
    accretors). The rotation rates and areal coverages are set at 5
    days and 10\% respectively (changing these has little effect). The
    top panels also include as a solid green line the 1 Myr pre-MS
    isochrone of \citet{siess_2000} adjusted to a distance modulus of
    7 and an extinction of $A_V=2$, simulating a reddened background
    population. The top panels then seperate the systems by accretion
    rate with \logmdot = $-$9, $-$10 and $-$11 \& $-$12 in the top left panel,
    and \logmdot = $-$6, $-$7 and $-$8 in the top right panel, shown blue,
    black and red dots respectively for both cases. The lower panels
    then split the systems into groups by inclination, with $\theta
    \leq$ 48$^{\circ}$, $\theta \geq$ 77$^{\circ}$ and $\theta=$ 56,
    64 \& 71$^{\circ}$, plotted as blue, red and black dots
    respectively.\label{age_mass_VVI}}
\end{figure*}

@@

If one adopts the range of input parameters we have used (see Section
\ref{par_space} for justification), our simulated photometry shows,
qualitatively, that a coeval 1 Myr population of accreting BD stars
and BDD systems, with typical accretion rates and range of
inclinations, will exhibit a significant scatter in apparent
isochronal age. Furthermore, objects with typical (and extreme)
accretion rates are scattered sufficiently in CMD space to prohibit
their identification as pre-MS BDs. Indeed, these objects would not be
included in a photometrically selected sample of BDs, and as such are
unlikely to be assigned the correct masses or ages. The scatter from
the naked 1 Myr systems, generally, increases with increasing
accretion rate. It is important to note at this point that these
conclusions are qualitative, and obviously based on our
assumptions. However, the repercussions for isochronal age derivation
and sample selection in the BD regime could be profound. Indeed this
study has only included the effects of current or ongoing accretion
from a disc, it has neglected any effects of accretion, both past and
present, on the evolution of the central star. This past accretion
could also act to reduce the stars radius, accelerating contraction
\citep{tout_1999,siess_1999} and introducing additional scatter in a
coeval population proportional to the range in accretion rates
\citep[see][for full discussion]{mayne_2008}. In addition we have
shown that the typical accretion rates may scatter BDD systems into
the region of a CMD occupied by the CTTS or background MS locus.

>>>>

The interpretation of the spectral energy distributions (SEDs) of
pre-main-sequence (PMS) stars involves trying to distinguish the
various contributions to the continuum from the hot spots at the base
of the accretion flow, the photospheric flux, and the near-IR flux
from the dusty inner disc. The complexity of the interplay between
these contributions is exacerbated by other, geometrical, effects such
as the inclination (which changes the projected area of the inner disc
wall visible to the observer) and the outer disc structure (which may
obscure the inner disc for high inclinations)
\citep{walker_2004,tannirkulam_2007}.

@@

Comparison of accretion rates across the pre-main-sequence (PMS) mass
spectrum has indicated that the accretion rate is strongly correlated
with PMS mass, \citep{muzerolle_2003,natta_2004,natta_2006}, with an
approximate form $\dot{M}_{\rm acc}\propto M^2_*$. Since in the
canonical picture the accretion rate is driven by the disc viscosity,
and should be independent of the mass of the central object, this
correlation is somewhat surprising. There is some danger that the
correlation is the result of, or at least strengthened by, the
presence of observational biases \citep{clarke_2006}. At the high-mass
(CTTS) end of the mass spectrum the lowest accretion-rates cannot be
measured via continuum methods since the excess is too small in
contrast with the photospheric emission. The emission lines will also
be weak, and indeed the H$\alpha$ equivalent width (EW) may be less
than the 10\AA\ that traditionally demarcates classical from
weak-lined T Tauri stars \citep[although this boundary may well be
spectral type dependent][]{navascues_2003}. Such objects should show
doppler-broadened profiles, but the line wings will be weak, and the
presence of underlying H$\alpha$ absorption may become dominant. Thus
there may well be a population of low-accretion-rate CTTS which are
current missing from the surveys.

There are also risks of observational biases at the low-mass end of
the correlation. Radiative-transfer modelling of H$\alpha$ emission
from accreting BDs requires high temperatures ($>$\, 10\,kK) in the
accretion funnels in order to recover the level of emission that is
observed \citep{muzerolle_2003,natta_2004,natta_2006}. Although
cooling rate arguments can be invoked to explain the presence of such
high temperatures, the line ratios of Pa$\beta$ and Br$\gamma$
indicate that the hydrogen emission may not arise in the funnel flows
\citep{gatti_2006}. However, recently some continuum measurements of
accreting BDs have been conducted \citep{herczeg_2009}, and these
broadly support the H$\alpha$ rates.

@@

In order to model an accreting BDD system we must first model the
underlying photospheric flux. We have adopted a BD stellar interior
and atmospheric model grid and have then constructed the total
photospheric flux for any input value of stellar age and mass by
interpolating for surface gravity ($\log g$), effective temperature
($T_{\rm eff}$), radius ($R_*/R_{\odot}$) and luminosity
($L_*/L_{\odot}$). These values were then used to interpolate
atmospheric spectra for flux (ergs s$^{-1}$cm$^{-2} {\rm \AA}^{-1}$)
from 1200 to 2$\times$10$^7$ \AA. The spectra were subsequently
resampled onto 200 logarithmically spaced points. Careful inspection
ensured that no spectral features were removed during resampling. The
stellar interior models used for this study are the `DUSTY00' models
of \cite{chabrier_2000} combined with the `AMES-Dusty', atmospheric
models of \cite{chabrier_2000}, which are all available
online\footnote{http://perso.ens-lyon.fr/france.allard/}. For our
$T_{\rm eff}$ range of $\approx $3000\,K$<T_{\rm eff}<$1600\,K the
AMES-Dusty atmospheres are the most applicable (2700\,K$>T_{\rm
  eff}>$1700\,K). We did try including dynamic application of
atmospheres based on the derived $T_{\rm eff}$, i.e. using AMES-Cond
for $T_{\rm eff}<$1700\,K, but this resulted in large discontinuities
between the model atmospheres and resulting spectra. Additionally, for
the higher temperatures in our range \cite{martin_2000} have found the
NextGen models to be more applicable for $\sim T_{\rm
  eff}>2300$K. However, as with the lower boundary this only applies
to the edge of our temperature range and we feel it is better to use a
consistent set of atmosphere and interior models across our
grid. Since this only affects stars at the very edge of our
temperature range, i.e. for the oldest and lowest mass objects (for
the AMES-Cond case), we have adopted the AMES-Dusty models throughout.

@@

Finally, a blackbody flux distribution is generated at $T_{\rm acc}$
and added onto the intrinsic stellar photospheric flux. In general one
would expect this to be an overestimate of the accretion flux, as for
BDs large convective zones are expected on the stellar surface, and
some of the accretion energy may act to further drive these convective
currents, meaning flux is lost. It is worth noting however that
observationally UV excesses are often used to recreate and then
subtract an assumed accretion flux using a blackbody flux curve, which
is essentially the reverse of this method.

@@

As our models assume magnetic-truncation at the co-rotation radius, we
have implicitly generated an inner hole. This also means that the disc
will have an inner wall at this radius. Evidence for inner walls in
circumstellar discs is apparent from the SEDs of disc systems, where a
peak in emission is found between 2 and 3\,$\mu$m. The temperatures
reached by such inner walls are expected to generate thermal flux
contributions within this wavelength range \citep{dullemond_2001}. The
circumstellar disc receives direct photospheric radiation and is
strongly heated. This leads to an increased scaleheight, or `puffing
up' of the inner disc rim, and subsequent shadowing of the region
immediately beyond it \citep{dullemond_2001}. The density-dependent
nature of the dust sublimation temperature \citep{pollack_1994} may
also play a role in shaping the inner rim
\citep{isella_2005,tannirkulam_2007}\footnote{Although we recognise
  that the inner hole in protostellar discs may be the result of giant
  planet formation or photoevaporation \citep{dahm_2009,najita_2007},
  these two mechanisms should result in a negligible mass-accretion
  rate \citep{dahm_2009} and therefore would represent a distinct
  evolutionary state of brown dwarf disc (BDD) systems to that of the
  CTTS-like phase we are considering here.}.

@@

Our treatment of dust sublimation is similar to that detailed in
\cite{tannirkulam_2007}, but with some enhancements to ensure a swift
convergence of the sublimated rim. The dust sublimation proceeds as
follows: An initial temperature distribution is found for the
optically-thin limit by setting the global dust-to-gas ratio to a tiny
value. Subsequent radiative equilibrium iterations are performed using
the adopted dust-to-gas ratio of 0.01, but limiting the maximum
optical depth across a given cell to $\tau_{\rm max}$. Cells whose
temperature exceeds the local dust sublimation temperature have their
dust-to-gas ratio set to zero. Radiative equilibrium iterations and
sublimation sweeps are performed at $\tau_{\max} = $0.1, 1, and 10,
with a final iteration of $\tau_{\max} = \infty$. Adequately solving
the radiative-equilibrium necessitates resolving the disc's effective
photosphere, and adaptive mesh refinement is used to split the grid at
the optically-thin/optically-thick boundary in order that the maximum
cell size at this boundary is $<1$ at a wavelength of 5500\,\AA. Once
a self-consistent sublimated rim has been determined the equation of
hydrostatic equilibrium is solved throughout the disc, and the
sublimation iterations are restarted (the change in the density
structure from the hydrostatical equilibrium step naturally feeds back
into the shape of the inner rim). Five hydrostatic equilibrium steps
are performed to ensure convergence, although a stable disc structure
is normally found after three such iterations. For a more
sophisticated, and chemically more complex, study of dust sublimation,
albeit for higher mass objects see \cite{kama_2009}.

@@

Accretion rates derived for pre-MS stars are of order \logmdot = $-6$
to $-11$ \citep{natta_2006}, with the largest accretion rates found in
so-called FU Orionis type objects. For the more typical accretion
rates \citep[$\dot{M}=$10$^{-11}$ to 10$^{-8.9} M_{\odot}yr^{-1}$, for
TTS,][]{dahm_2009}, several studies have now suggested that the
accretion rate is strongly correlated with the mass of the central
star. This relationship was perhaps first suggested by
\cite{muzerolle_2003} using various accretion diagnostics. Later,
\citep{muzerolle_2005} derived a relationship of approximately
$\dot{M}\propto M_*^2$. Further evidence was put forward by
\cite{natta_2004}, where accretion rates as low as 5$\times$ 10$^{-12}
{\rm M}_{\odot} {\rm yr}^{-1}$ were found for BD stars. More recently,
even lower accretion rates of $\approx$10$^{-13} {\rm M}_{\odot} {\rm
  yr}^{-1}$, have been derived for BDs by \cite{herczeg_2009}.
Further support for a dependence of accretion rate on stellar mass was
apparent in the significantly more homogeneous dataset of
\cite{natta_2006}. \cite{natta_2006} analysed a set of accretion rates
and masses derived for BD stars in $\rho$ Ophiuchi and compared these
results to stars in Taurus. They found that the accretion rate scales
with central object mass into the BD regime, although with significant
scatter.

@@

For further analysis, and throughout the rest of this paper, we will
separate the systems into two groups. These groups are those systems
with \logmdot$\leq-9$ and \logmdot$\geq-9$, classed as typical and
extreme accretors. This distinction is based on the point at which the
systems undergo both significant sublimation ($\Delta R_{\rm
  inner}>1R_*$) and have an accretion luminosity that is comparable to
the photospheric luminosity (see Section~\ref{acc_dominance}). In
order to examine the effect of dust sublimation on our models we have
measured the radius of the inner edge of the dust disc by integrating
from the centre along the midplane until an optical depth of 2/3 (at
5500\AA) is reached. Figure \ref{inner_temp} shows the radial density
distribution of the disc.  In this case the the final inner edge
radius and therefore temperature is no longer strongly correlated with
the rotation rate. Figure \ref{inner_temp} shows the final inner edge
location ($R_{\rm inner}$, $R_*$) against the temperature of the inner
wall ($T_{\rm inner}$, K). The left panel shows the systems designated
as typical accretors and the right panel those with extreme accretion
rates.  For both panels the systems with rotation periods of 0.5 and 5
days are plotted in red and blue respectively. Those systems where the
change in inner radius was greater than 1$R_*$, and therefore classes
as significantly sublimated, are plotted as crosses.

\begin{figure*}
\includegraphics[scale=0.5,angle=90]{./Fig/inner_temp.ps}
   \caption{Figure showing data for the inner edge location ($R_{\rm
       inner}$, $R_*$). The typical accretors are shown in the
     left panel and the extreme accretors in the
     right panel (see text for explanation). In both
       panels separate the systems by period, with those systems with
     rotation periods of 0.5 and 5 days shown as red and blue symbols
     respectively. In addition, systems where $\Delta R_{\rm inner} <
     1R_*$ are plotted as crosses and $\Delta R_{\rm inner} > 1R_*$ as
     filled circles (this is only achieved by some systems classed as
     extreme accretors).\label{inner_temp}}
\end{figure*}

@@

For our model grid the inner edge shape clearly changes with
increasing flux levels and therefore accretion rate. This, for
individual systems will lead to the reduced dependence of the IR
colours on inclination as the edge moves from a vertical to a convex
boundary as found by \cite{isella_2005}. Practically, for populations
of stars however disentangling this effect from the influences of
various other parameters (such as changing disc scaleheights and
accretion rates) on the IR colours is difficult. Essentially, matching
flux levels between two systems whilst maintaining a difference in
inner wall shape is impossible.

Therefore, for individual systems changes in the accretion rate and
disc structure lead to significant changes in magnitudes and
colours. The data presented in this section are for an isolated mass
and age. However, the trends described are evident throughout our
model grid for any given subset, and as such are representative. The
scatter or changes in magnitude in the individual systems will change
as a function of the remaining variables, but the dominant input
variables affecting simulated photometry are accretion rate and
inclination. For populations of stars these changes in magnitude and
colour act to scatter or spread a BD locus in photometric space.

@@

Practically, most parameters for young pre-MS and BD stars are derived
from surveys of populations, usually open clusters, using broadband
photometry and subsequently constructed colour-magnitude and
colour-colour diagrams (CMDs and CoCoDs respectively, hereafter).  In
this Section we explore the consequences of our model grid on the
derivation of the primary parameters of age, mass and disc fractions,
from populations. This in turn leads to highlighting selection effects
with, for instance, the mass to accretion rate relation.

@@

As discussed in Section \ref{disc_struct} the inner edge location is
correlated with inner edge temperature (albeit differently for the
typical and extreme accretors). As noted by \cite{meyer_1997} this
could lead to a correlation of IR excess with inner edge position. As
expected however, scatter caused by variations in inclination and disc
structure act to remove this correlation for populations.  The
correlation for an individual set of a systems, i.e. all variables
fixed except rotation rate, between IR colour and rotation rate is
preserved. However, as shown in Section \ref{disc_struct} the changes
caused by accretion rate and inclination angle are the most
significant, and act to remove any correlation between rotation rate
and IR colours for populations. In fact, for our grid the simulated
photometry shows no significant correlation between rotation rate and
IR colours. The work of \cite{meyer_1997} studied flat accreting disc,
therefore, as we have increased variation in disc structure by
applying vertical hydrostatic equilibrium it would be interesting to
see if any correlation appears for analytically defined disc
structures. We have run a set of parallel models using analytical
disc structures and will publish the results in a subsequent
publication. Overall, as found in \cite{walker_2004}, the dominant
scattering effect for BD disc systems in an optical CMD appears to be
caused by accretion rate and inclination. Therefore for derivation of
parameters we explore the scatter in BD locii as a function of
accretion rate and inclination.

@@

The use of pre-MS and BD isochrones for the derivation of single star
parameters is at present not proven to be reliable \citep[see
discussion in][]{mayne_2008}. Practically, therefore, median ages are
derived from populations. Subsequently, derived masses are still
unreliable but at least based on a consistent age. This problem is
being addressed by Bell et al (in prep), where $K$ band photometry and
known eclipsing binaries are being used to refine pre-MS
isochrones. In this section we plot the data for our 1 Myr systems
only and explore the resulting scatters caused by the disc presence
and accretion luminosity.

Figures \ref{age_mass_VVI} and \ref{age_mass_JJK} shows CMDs in $M_V$
\& $(V-I)_0$ and $M_J$ \& $(J-K)_0$ respectively. The left panels of
both figures shows stars classed as typical accretors with accretion
rates of \logmdot = $-$9, $-$10 and $-$11 \& $-$12, shown as blue,
black and red dots respectively in the top panels. The right panels of
Figures \ref{age_mass_VVI} and \ref{age_mass_JJK} show systems with
extreme accretion rates, with \logmdot = $-$6, $-$7 and $-$8 shown as
blue, black and red dots respectively, in the top right panel. The
bottom panels then show the systems separated into groups by
inclination. These groups are $\theta \leq 48^{\circ}$ as blue dots
(classed as face-on systems), $\theta> 56$ \& $64^{\circ}$ (classed as
the expected systems, as the expectation value of
$cos(\theta)=60^{\circ}$) as black dots and $\theta \geq 71^{\circ}$
(classed as edge-on systems) as red dots (for typical, bottom left,
and extreme, bottom right accretion rates). The left panels then show
naked stars isochrones (created from our grid of simulated photometry)
for 1 Myrs at accretion rates of \logmdot = $-$12 and $-$9, as solid
and dashed lines respectively (with coverage of 10\% and rotation
rates of 5 days). Whereas the left hand panels show naked stars
isochrones of \logmdot = $-$6 and $-$9 as solid and dashed black lines
respectively (with coverage of 10\% and rotation rates of 5 days. The
solid green line, in the top panels, shows the 1 Myr isochrone of
\cite{siess_2000} adjusted to a distance of 250 pc and an extinction of
$A_V=2$ mag, simulating a background population of CTTS stars.

\begin{figure*}
  \centering
  \includegraphics[scale=0.6,angle=90]{./Fig/age_mass_VVI.ps}
  \caption{Figure showing CMDs in $M_V$, $(V-I)_0$ for typical
    accretors (\logmdot $\leq$ $-$9) in the left panels, and extreme
    accretors (\logmdot $>-9$) in the right panels. The black dashed
    and solid lines correspond to naked BD isochrones (from our grid)
    with \logmdot\ $-$9 and $-$12 for the left panels (typical
    accretors) and -9 and -6 in the right panels (extreme
    accretors). The rotation rates and areal coverages are set at 5
    days and 10\% respectively (changing these has little effect). The
    top panels also include as a solid green line the 1 Myr pre-MS
    isochrone of \citet{siess_2000} adjusted to a distance modulus of
    7 and an extinction of $A_V=2$, simulating a reddened background
    population. The top panels then separate the systems by accretion
    rate with \logmdot = $-$9, $-$10 and $-$11 \& $-$12 in the top left panel,
    and \logmdot = $-$6, $-$7 and $-$8 in the top right panel, shown blue,
    black and red dots respectively for both cases. The lower panels
    then split the systems into groups by inclination, with $\theta
    \leq$ 48$^{\circ}$, $\theta \geq$ 77$^{\circ}$ and $\theta=$ 56,
    64 \& 71$^{\circ}$, plotted as blue, red and black dots
    respectively.\label{age_mass_VVI}}
\end{figure*}

@@

If one adopts the range of input parameters we have used (see Section
\ref{par_space} for justification), our simulated photometry shows,
qualitatively, that a coeval 1 Myr population of accreting BD stars
and BDD systems, with typical accretion rates and range of
inclinations, will exhibit a significant scatter in apparent
isochronal age. Furthermore, objects with typical (and extreme)
accretion rates are scattered sufficiently in CMD space to prohibit
their identification as BDs. Indeed, these objects would not be
included in a photometrically selected sample of BDs, and as such are
unlikely to be assigned the correct masses or ages. The scatter from
the naked 1 Myr systems, generally, increases with increasing
accretion rate. It is important to note at this point that these
conclusions are qualitative, and obviously based on our
assumptions. However, the repercussions for isochronal age derivation
and sample selection in the BD regime could be profound. Indeed this
study has only included the effects of current or ongoing accretion
from a disc, it has neglected any effects of accretion, both past and
present, on the evolution of the central star. This past accretion
could also act to reduce the stars radius, accelerating contraction
\citep{tout_1999,siess_1999} and introducing additional scatter in a
coeval population proportional to the range in accretion rates
\citep[see][for full discussion]{mayne_2008}. In addition we have
shown that the typical accretion rates may scatter BDD systems into
the region of a CMD occupied by the CTTS or background MS locus.
